Calibrating seed-based alignment heuristics with Sesame

### 2020-03-20 manuscript.tex
Add detail for off-traget skip seeds

With these notations, he shortest $\Downarrow_j$-segment consists of $x$
symbols $\square$ followed by the $\Downarrow_j$ terminator, so it has
weighted generating function $(qz)^xpz$. The other segments have a
multiple of $n+1$ extra $\square$ symbols so their weighted
generating functions are $(qz)^{x+(n+1)}pz, (qz)^{x+2(n+1)}pz, \ldots,
(qz)^{x+m(n+1)}pz$. Summing over those cases, we finally obtain
\begin{eqnarray}
\label{eq_H}
H_{i,j}(z) &=& (qz)^xpz \cdot \big( 1 + (qz)^{n+1} +
  \ldots + (qz)^{m(n+1)} \big), \\
\notag
  &\;& \text{where } x = i-j-1 \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma+i-1-x}{n+1} \right\rfloor.
\end{eqnarray}

@@

$P(\overline{S}_0 \cap \overline{S}_1)$ is the probability that the read
has no seed for the target or for the first duplicate---numbering is
arbitrary here, the first duplicate can be any fixed duplicate. As
illustrated in section~\ref{sec:symbolic}, we first recode the reads using
a specialized alphabet to simplify the problem.

It will be useful to consider a more general problem where we
have two sequences of interest labelled $+$ and $-$. The $+$ sequence
stands for the target and that the $-$ sequence stands for its duplicate.
We then define the dual alphabet $\tilde{\mathcal{A}}_0 = \{\square, |,
\downarrow_{/1}^-, \downarrow_{/2}^-, \ldots, \downarrow_{/1}^+,
\downarrow_{/2}^+, \ldots, \Downarrow\}$. The symbols $\downarrow_{/j}^-$
($j \geq 1$) signify that the nucleotide is a mismatch against the $-$
sequence only, the symbols $\downarrow_{/j}^+$ ($j \geq 1$) signify that
it is a mismatch against the $+$ sequence only, and the symbol
$\Downarrow$ signifies that it is a mismatch against both. As before,
every other nucleotide is replaced by the symbol $\square$, and the
terminator $|$ is appended to the end of the read. We again define reads
as sequences of segments (zero or more $\square$ symbols followed by a
terminator), except that now the terminators are the symbols
$\downarrow_{/j}^-$, $\downarrow_{/j}^+$ and $\Downarrow$ ($j \geq 1$).
The tail, as usual, is terminated by the symbol $|$.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_dual.pdf}
\caption{\textbf{Example of dual encoding.}
An example of read is represented in the dual alphabet. The symbols
$\downarrow_{/j}^-$ ($j \geq 1$) represent a mismatch against the $-$
sequence, the symbols $\downarrow_{/j}^+$ ($i \geq 1$) represent a
mismatch against the $+$ sequence, and the symbol $\Downarrow$ represents
a mismatch against both. The index $i$ is the match length of the sequence
that is not mismatched. The symbolic $+$ and $-$ sequences are represented
below, where an open square stands for a match and a closed square stands
for a mismatch.}
\label{fig:dual}
\end{figure}

@@

With these definitions, $a$, $b$, $c$ and $d$ are the probabilities of
occurrence of the symbols $\square$, $\downarrow_{/j}^-$,
$\downarrow_{/j}^+$ and $\Downarrow$, respectively ($j \geq 1$). The next
sections clarify how this is used to compute the weighted gerating
functions of interest.

@@

If the next terminator is $\downarrow_{/i}^+$, there is a match of size
$i$ for the $-$ sequence, so the segment contains $i-1$ symbols
$\square$ (one of the matches is accounted for by the terminator). The
weighted generating function is thus $(az)^{i-1}cz$. Likewise, if the next
terminator is $\downarrow_{/i}^-$, the weighted generating function of the
segment is $(az)^{i-1}bz$. The terminators $\downarrow_{/i}^+$ and
$\downarrow_{/i}^-$ are disallowed for $j \geq \gamma$ because this would
create a match of size $\gamma$ or greater for one of the sequences.

\subsection{Segments following $\downarrow_{/j}^+$}

At a $\downarrow_{/j}^+$ terminator, the match length of the $+$ sequence
is 0 and the match length of the $-$ sequence is $j$. The next segment can
thus have up to $\gamma-1$ matches for the $+$ sequence, but only
$\gamma-j-1$ matches for the $-$ sequence, lowering the maximum size of
the segment. If the next terminator is $\Downarrow$ or $|$, the segments
have weighted generating functions $(1 + az + \ldots +
(az)^{\gamma-j-1})dz$ and $1 + az + \ldots + (az)^{\gamma-j-1}$,
respectively.

If the next terminator is $\downarrow_{/i}^-$, the weighted generating
function is $(az)^{i-1}bz$ as in the previous section. The difference is
that the terminators $\downarrow_{/i}^-$ are allowed only for $1 \leq i
\leq \gamma-j$, otherwise this would create a match of size $\gamma$ or
greater for the $-$ sequence.

If the next terminator is $\downarrow_{/i}^+$, the situation is slightly
more complicated, because this imposes $j < i \leq \gamma-1$. Indeed,
there were already $j$ matches for the $-$ sequence at the terminator
$\downarrow_{/j}^+$, and there will be more at the end of the segment
because it has no mismatch for the $-$ sequence. Taking this into account,
we see that the weighted generating function of those segments is
$(az)^{i-j-1}cz$.

\subsection{Segments following $\downarrow_{/j}^-$}

We can find the weighted generating functions by just reversing the $+$
and $-$ signs in the previous section. This way, we see that the weighted
generating function of the segments terminated by $\Downarrow$ or $|$ are
$(1 + az + \ldots + (az)^{\gamma-j-1})dz$ and $1 + az + \ldots +
(az)^{\gamma-j-1}$, respectively.

Likewise, the weighted generating function of the segments terminated by
$\downarrow_{/i}^+$ is $(az)^{i-1}cz$, where $1 \leq i \leq \gamma-j$; and
the weighted generating function of the segments terminated by
$\downarrow_{/i}^-$ is $(az)^{i-j-1}bz$, where $j < i \leq \gamma-1$.

@@

We now have all the elements to specify the transfer matrix or reads with
no seed for either sequence. Recall that $a$ is the probability of a
double match, $b$ is the probability of a mismatch only against the $-$
sequence, $c$ is the probability of a mismatch only against the $+$
sequence and $d$ is the probability of a double mismatch. For notational
convenience, we define
\begin{equation}
% These expressions are not needed for reference in the text, but
% they are still numbered so that the code can reference to it.
\begin{gathered}
r_i^+(z) = (az)^i cz, \\
r_i^-(z) = (az)^i bz, \\
R_i(z) = \big( 1 + az + \ldots + (az)^i \big) dz, \\
F_i(z) = 1 + az + \ldots + (az)^i.
\end{gathered}
\end{equation}

@@

where $\tilde{A}(z)$, $\tilde{B}_0(z)$, $\tilde{C}_0(z)$ and
$\tilde{D}(z)$ are matrices of dimensions $(\gamma-1) \times (\gamma-1)$
that are defined as

@@

As before, the term of interest is the top right entry of
$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. To see why, observe that
every read can be prepended by $\Downarrow$-segments and only by those
(every other terminator would imply that one of the two sequences has a
nonzero match size at the start of the read). Thus reads are precisely the
sequences of segments that can follow the symbol $\Downarrow$ and that are
terminated by a tail, the weighted generating function of which is located
in the top right entry of the matrix.

$\tilde{M}_0(z)$ is too complex to compute a closed expression of
$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. It is easier to proceed as
in section~\ref{sec:example_skip} and to compute the powers of
$\tilde{M}_0(z)$ up to a finite value. This is done once again using the
arithmetic of truncated polynomials. Since each segment except the tail
contains a mismatch against at least one sequence, we define $\tilde{p}$
as the upper bound on the probability of a mismatch, \textit{i.e.}
$\tilde{p} = \max\{b,c,d\}$. The updated formula (\ref{eq:bound}) now
gives an upper bound of the probability that a read of size $k$ contains
$s$ or more mismatches as
\begin{equation*}
Pr(X \geq s) \leq \exp \left( (s-k)\log \frac{k-s}{k(1-\tilde{p})} -s\log
\frac{s}{k\tilde{p}} \right).
\end{equation*}

@@

With these values, we can fully specify the matrix $\tilde{M}_0(z)$, and
compute its powers until the right-hand side of the expression above is
lower than a set fraction $\varepsilon$ of the current value of $a_k$,
which gives an accurate estimate of $P(\overline{S}_0 \cap
\overline{S}_1)$.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.65]{curves_exact.pdf}
\caption{\textbf{Off-target seeding probabilities (exact seeds).} The
surfaces show the probability that seeding is off target for exact seeds
of size $\gamma=19$ in reads of size $k=50$ nucleotides sequenced with an
error rate $p=0.01$. The divergence rate $\mu$, defined as the probability
that a given duplicate differs from the target at any given position. Note
the factor 1000 in the scale of the z-axis.} \label{fig:curves_exact}
\end{figure}

Figure~\ref{fig:curves_exact} shows the result for a number of duplicates
$N$ from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The first
observation is that the probability that seeding is off target increases
with $N$. This is also clearly from expression (\ref{eq:Poff}). This can
also be understood intuitively because the probability of null seeding
decreases when there are more potential candidates, and since the
probability of on-target seeding does not change, the probability of
off-target seeding must increase.

The second observation is that there exists a ``worst'' value of $\mu$
situated around 0.07. When $\mu$ is much smaller, the duplicates tend to
be exactly identical to the target, making it impossible that there is a
seed for a duplicate but no seed for the target. When $\mu$ is much
larger, the duplicates are far from the target and they are unlikely to be
in the candidate set. In expression (\ref{eq:Poff}), the only term that
depends on $\mu$ is $P(\overline{S}_0 \cap \overline{S}_1)$, and it is
clear that the minimum of expression (\ref{eq:Poff}) corresponds to the
maximum of $P(\overline{S}_0 \cap \overline{S}_1)$. This is why the worst
value of $\mu$ is the same for every $N$.

@@

Again we recode the reads in a specialized alphabet. We define the
skip-$n$ dual alphabet as $\tilde{\mathcal{A}}_n = \{\square, *, |,
\Downarrow_0, \Downarrow_1, \Downarrow_2, \ldots, \Downarrow_n,
\downarrow^-_{/1}, \downarrow^-_{/2}, \ldots, \downarrow^+_{/1},
\downarrow^+_{/2}, \ldots\}$. The symbols $\square$, $|$,
$\downarrow^-_{/j}$ and $\downarrow^+_{/j}$ ($j \geq 1$) have the same
meaning as in the dual alphabet of section~\ref{sec:dual}. The symbols
$\Downarrow_j$ ($0 \leq j \leq n$) indicate that both sequences have match
length 0 and that the next non-skipped position is $j$ nucleotides
further. The new symbol $*$ indicates that the nature of the nucleotide
does not matter, as we will explain below.

@@

Also observe that the symbol $\Downarrow_1$ is followed by a $*$ to
indicate that this nucleotide has no influence on the presence of a seed,
regardless 
. The reason is that there
is a double mismatch at the $\Downarrow_1$ symbol so there can be no seed.
But since the next potential seed starts only at the next tick mark,
\textit{i.e.} one nucleotide further downstream,

@@

As in the previous section, let $a$, $b$, $c$ and $d$ denote the
probabilities of a double match, a mismatch against $-$, a mismatch
against $+$ and a double mismatch, respectively. This also corresponds to
the probabilities of occurrence of the symbols $\square$,
$\downarrow_{/j}^+$, $\downarrow_{/j}^-$ ($j \geq 1$) and $\Downarrow_j$
($0 \leq j \leq n$), respectively. With this information, we compute the
weighted generating functions of interest in the sections below.

\subsection{Segments following $\Downarrow_j$ ($1 \leq j \leq n)$}

After a $\Downarrow_j$ terminator, the match counter for both sequences is
at zero and the next potential seed starts only $j \geq 1$ nucleotides
further downstream. This means that the next $j$ nucleotides are just
filler so it does not matter which sequence they match or mismatch.

There are only two scenarios to consider: if the end of the read is within
the next $j$ nucleotides, then

@@

Hello world!

@@

As in the case of the skip encoding, a fair amount of bookkeeping is
required to specify the weighted generating functions of interest.
We reuse the quantities $a+b+c+d=1$ and give them the same meaning as in
section~\ref{sec:dual}. For notational convenience, we define
\begin{eqnarray}
\label{eq_N}
N_i(z) &=& 1+z+\ldots+z^i, \\
W_j(z) &=& dz \cdot (az)^x \cdot \big( 1 + (az)^{n+1} + \ldots +
  (az)^{(n+1)m} \big), \\
  &\quad& \text{where } x = n-j \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma-1-x}{n+1} \right\rfloor. \\
U_{i,j}(z) &=& dz \cdot (az)^x \cdot
  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big) +
\begin{cases}
  bz\cdot(az)^x \text{ if } j < y, \\
  0 \text{ otherwise}
\end{cases} \\
  &\quad& \text{where } x = -i-j-1 \modulo{n+1},
  y = n-i+1 \modulo{n+1}, \\
  &\quad& \text{and } m = \left\lfloor \frac{\gamma-1-i-x}{n+1}
    \right\rfloor, \\
V_{i,j}(z) &=& dz \cdot (az)^x \cdot
  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big) +
\begin{cases}
  cz\cdot(az)^x \text{ if } j \leq x, \\
  0 \text{ otherwise}
\end{cases} \\
  &\quad& \text{ where } x = -i-j-1 \modulo{n+1},
  m = \left\lfloor \frac{\gamma-1-i-x}{n+1} \right\rfloor.
\end{eqnarray}

>>>>

With these notations, he shortest $\Downarrow_j$-segment consists of $x$
symbols $\square$ followed by the $\Downarrow_j$ terminator, so it has
weighted generating function $(qz)^xpz$. The other segments have a
multiple of $n+1$ extra $\square$ symbols so their weighted
generating functions are $(qz)^{x+(n+1)}pz, (qz)^{x+2(n+1)}pz, \ldots,
(qz)^{x+m(n+1)}pz$. Summing over those cases, we finally obtain
\begin{eqnarray}
\label{eq_H}
H_{i,j}(z) &=& (qz)^x \big( 1 + (qz)^{n+1} +
  \ldots + (qz)^{m(n+1)} \big)pz, \\
\notag
  &\;& \text{where } x = i-j-1 \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma+i-1-x}{n+1} \right\rfloor.
\end{eqnarray}

@@

$P(\overline{S}_0 \cap \overline{S}_1)$ is the probability that the read
has no seed for the target or for the first duplicate---numbering is
arbitrary here, the first duplicate can be any fixed duplicate. As in
section~\ref{sec:symbolic}, we first recode the reads using a specialized
alphabet to simplify the problem.

It will be useful to consider a more general problem where we
have two sequences of interest labelled $+$ and $-$. The $+$ sequence
stands for the target and that the $-$ sequence stands for its duplicate.
We then define the dual alphabet $\tilde{\mathcal{A}}_0 = \{\square, |,
\downarrow_{/1}^-, \downarrow_{/2}^-, \ldots, \downarrow_{/1}^+,
\downarrow_{/2}^+, \ldots, \Downarrow\}$. The symbols $\downarrow_{/j}^-$
($j \geq 1$) signify that the nucleotide is a mismatch against the $-$
sequence only, the symbols $\downarrow_{/j}^+$ ($j \geq 1$) signify that
it is a mismatch against the $+$ sequence only, and the symbol
$\Downarrow$ signifies that it is a mismatch against both. As before,
every other nucleotide is replaced by the symbol $\square$, and the
terminator $|$ is appended to the end of the read. We again define reads
as sequences of segments (zero or more $\square$ symbols followed by a
terminator), except that now the terminators are the symbols
$\downarrow_{/j}^-$, $\downarrow_{/j}^+$ ($j \geq 1$) and $\Downarrow$.
The tail, as usual, is terminated by the symbol $|$.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_dual.pdf}
\caption{\textbf{Example of dual encoding.}
An example of read is represented in the dual alphabet. The symbols
$\downarrow_{/j}^-$ ($j \geq 1$) represent a mismatch against the $-$
sequence, the symbols $\downarrow_{/j}^+$ ($j \geq 1$) represent a
mismatch against the $+$ sequence, and the symbol $\Downarrow$ represents
a mismatch against both. The index $i$ is the match length of the sequence
that is not mismatched. The symbolic $+$ and $-$ sequences are represented
below, where an open square stands for a match and a closed square stands
for a mismatch.}
\label{fig:dual}
\end{figure}

@@

With these definitions, the weighted generating functions of the symbols
$\square$, $\downarrow_{/j}^-$, $\downarrow_{/j}^+$ ($j \geq 1$) and
$\Downarrow$ are $az$, $bz$, $cz$ and $dz$, respectively. The next
sections clarify how this is used to compute the weighted generating
functions of interest.

@@

If the next terminator is $\downarrow_{/j}^-$, there is a match of length
$j$ for the $+$ sequence, so the segment contains $j-1$ symbols $\square$
plus the terminator (which also matches the $+$ sequence). The weighted
generating function is thus $(az)^{j-1}bz$. By the same rationale, if the
next terminator is $\downarrow_{/j}^+$, the weighted generating function
of the segment is $(az)^{j-1}cz$.

The terminators $\downarrow_{/j}^+$ and $\downarrow_{/j}^-$ are disallowed
for $j \geq \gamma$ because this would create a seed for at least one of
the sequences.

\subsection{Segments following $\downarrow_{/i}^+$}

At a $\downarrow_{/i}^+$ terminator, the match length of the $+$ sequence
is 0 and the match length of the $-$ sequence is $i$. The next segment can
thus have up to $\gamma-1$ matches for the $+$ sequence, but only
$\gamma-i-1$ matches for the $-$ sequence, lowering the maximum size of
the segment. If the next terminator is $\Downarrow$ or $|$, the segments
have weighted generating functions $(1 + az + \ldots +
(az)^{\gamma-j-1})dz$ and $1 + az + \ldots + (az)^{\gamma-j-1}$,
respectively.

If the next terminator is $\downarrow_{/j}^-$, the weighted generating
function is $(az)^{j-1}bz$ as in the previous section. The difference is
that the terminators $\downarrow_{/j}^-$ are allowed only for $1 \leq j
\leq \gamma-i$, otherwise this would create seed for the $-$ sequence.

If the next terminator is $\downarrow_{/j}^+$, the situation is slightly
more complex because this imposes $i < j \leq \gamma-1$. Indeed,
there were already $i$ matches for the $-$ sequence at the terminator
$\downarrow_{/i}^+$, and there will be more at the end of the following
segment because it has no mismatch for the $-$ sequence. Taking this into
account, we see that the weighted generating function of those segments is
$(az)^{j-i-1}cz$ with $i < j \leq \gamma-1$.

\subsection{Segments following $\downarrow_{/i}^-$}

We can find the weighted generating functions by just reversing the $+$
and $-$ signs in the previous section. This way, we see that the weighted
generating function of the segments terminated by $\Downarrow$ or $|$ are
$(1 + az + \ldots + (az)^{\gamma-i-1})dz$ and $1 + az + \ldots +
(az)^{\gamma-i-1}$, respectively.

Likewise, the weighted generating function of the segments terminated by
$\downarrow_{/j}^+$ is $(az)^{j-1}cz$, where $1 \leq j \leq \gamma-i$; and
the weighted generating function of the segments terminated by
$\downarrow_{/j}^-$ is $(az)^{j-i-1}bz$ where $i < j \leq \gamma-1$.

@@

We now have all the elements to specify the transfer matrix of reads with
no seed for either sequence. Recall that $a$ is the probability of a
double match, $b$ is the probability of a mismatch only against the $-$
sequence, $c$ is the probability of a mismatch only against the $+$
sequence and $d$ is the probability of a double mismatch. For notational
convenience, we define
\begin{equation}
% These expressions are not needed for reference in the text, but
% they are still numbered so that the code can reference to it.
\begin{gathered}
r_i^+(z) = (az)^i cz, \\
r_i^-(z) = (az)^i bz, \\
R_i(z) = \big( 1 + az + \ldots + (az)^i \big) dz, \\
F_i(z) = 1 + az + \ldots + (az)^i.
\end{gathered}
\end{equation}

@@

where $\gamma$ is the minimum seed length, and where $\tilde{A}(z)$,
$\tilde{B}_0(z)$, $\tilde{C}_0(z)$ and $\tilde{D}(z)$ are matrices of
dimensions $(\gamma-1) \times (\gamma-1)$ that are defined as

@@

As before, the term of interest is the top right entry of
$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. To see why, observe that
every read can be prepended by $\Downarrow$-segments and only by those
(every other terminator would imply that one of the two sequences has a
nonzero match size at the start of the read). Thus reads are precisely the
sequences of segments that can follow the symbol $\Downarrow$ and that are
terminated by a tail, the weighted generating function of which is the top
right entry of the matrix.

$\tilde{M}_0(z)$ is too complex to compute a closed expression of
$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. It is easier to proceed as
in section~\ref{sec:example_skip} and to compute the powers of
$\tilde{M}_0(z)$ up to a finite value. This is done once again using the
arithmetic of truncated polynomials. Since each segment except the tail
contains a mismatch against at least one sequence, the top right entry of
$\tilde{M}_0(z)^{s+1}$ is the weighted generating function of reads that
contain $s$ mismatches (against any of the two sequences). We thus define
$\tilde{p}$ as the upper bound on the probability of a mismatch,
\textit{i.e.} $\tilde{p} = \max\{b,c,d\}$. The updated formula
(\ref{eq:bound}) now gives an upper bound of the probability that a read
of size $k$ contains $s$ or more mismatches as
\begin{equation*}
Pr(X \geq s) \leq \exp \left( (s-k)\log \frac{k-s}{k(1-\tilde{p})} -s\log
\frac{s}{k\tilde{p}} \right).
\end{equation*}

With this upper bound, we can compute the terms of the matrix partial sums
$\tilde{M}_0(z) + \tilde{M}_0(z)^2 + \ldots + \tilde{M}_0(z)^s$ until the
neglected terms become negligible, \textit{i.e.} until we can be sure that
the coefficient of interest $a_k$ is accurate to within chosen
$\varepsilon$.

@@

With these values, we can fully specify the matrix $\tilde{M}_0(z)$, and
compute its powers until the estimate of the coefficient of interest $a_k$
is accurate enough, finally giving a numerical value for $P(\overline{S}_0
\cap \overline{S}_1)$.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.65]{curves_exact.pdf}
\caption{\textbf{Off-target seeding probabilities (exact seeds).} The
surfaces show the probability that seeding is off target as a function of
the divergence rate $\mu$ and of the number $N$ of duplicates. The results
are shown for exact seeds of size $\gamma=19$ and for reads of size $k=50$
nucleotides sequenced with an error rate $p=0.01$. The divergence rate
$\mu$ is defined as the probability that a given duplicate differs from
the target at any given position. Note the factor 1000 in the scale of the
z-axis.}
\label{fig:curves_exact}
\end{figure}

Figure~\ref{fig:curves_exact} shows the result for a number of duplicates
$N$ from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The first
observation is that the probability that seeding is off target increases
with $N$. This is also clear from expression (\ref{eq:Poff}). This can
also be understood intuitively because the probability of not seeding the
target is fixed and the probability of having an empty candidate set
decreases as $N$ increases. As a result the probability that the candidate
set contains only invalid candidates increases with $N$.

The second observation is that there exists a ``worst'' value of $\mu$
situated around 0.07. When $\mu$ is much smaller, the duplicates tend to
be exactly identical to the target, making it impossible that there is a
seed for a duplicate but no seed for the target. When $\mu$ is much
larger, the duplicates are far from the target and they are unlikely to be
in the candidate set at all. In expression (\ref{eq:Poff}), the only term
that depends on $\mu$ is $P(\overline{S}_0 \cap \overline{S}_1)$, and it
is clear that the minimum of expression (\ref{eq:Poff}) corresponds to the
maximum of $P(\overline{S}_0 \cap \overline{S}_1)$. This is why the worst
value of $\mu$ is the same for every $N$.

@@

As before, we recode the reads in a specialized alphabet. We define the
skip-$n$ dual alphabet as $\tilde{\mathcal{A}}_n = \{\square, *, |,
\Downarrow_0, \Downarrow_1, \Downarrow_2, \ldots, \Downarrow_n,
\downarrow^-_{/1}, \downarrow^-_{/2}, \ldots, \downarrow^+_{/1},
\downarrow^+_{/2}, \ldots\}$. The symbols $\square$, $|$,
$\downarrow^-_{/j}$ and $\downarrow^+_{/j}$ ($j \geq 1$) have the same
meaning as in the dual alphabet of section~\ref{sec:dual}, \textit{i.e.}
the $\square$ symbol stands for a double match, the $|$ terminator marks
the end of the read, the $\downarrow^-_{/j}$ symbol indicates a mismatch
against the $-$ sequence with a match length of size $j$ for the $+$
sequence, and conversely the $\downarrow^+_{/j}$ symbol indicates a
mismatch against the $+$ sequence with a match of length $j$ for the $-$
sequence. The symbols $\Downarrow_j$ ($0 \leq j \leq n$) indicate that
both sequences have match length 0 and that the next non-skipped position
is $j$ nucleotides further. The symbol $*$ indicates that it does not
matter whether the nucleotide is a match or a mismatch, as we will explain
below.

@@

@@

Also observe that the symbol $\Downarrow_1$ is followed by a $*$ symbol,
indicating that it does not matter whether the nucleotide is a match or a
mismatch for any of the two sequences. After the $\Downarrow_1$ symbol,
both sequences have match length 0 and one has to ``wait'' for a potential
new seed one nucleotide further downstream. As in appendix~\ref{app_Mn},
the $\Downarrow_j$ symbols are followed by $j$ symbols $*$, unless the end
of the read comes before the next non-skipped position. After a sequence
of $*$ symbols, the read is either finished or at a non-skipped position.
In the second case, the is in the same state as after a $\Downarrow_0$
symbol, so we can consider that the $*$ symbol allows the terminators
$\Downarrow_j$ ($1 \leq j \leq n$) to just ``fast forward'' to either
$\Downarrow_0$ or $|$, as shown in appendix~\ref{app_Mn}.

As in the previous section, let $a$, $b$, $c$ and $d$ denote the
probabilities of a double match, a mismatch against $-$, a mismatch
against $+$ and a double mismatch, respectively. With these definitions,
the weighted generating functions of the symbols $\square$,
$\downarrow_{/j}^-$ and $\downarrow_{/j}^+$ ($j \geq 1$) are $az$, $bz$
and $cz$, respectively. The weighted generating function of the $*$ symbol
is $z$, and that of the symbols $\Downarrow_j$ ($0 \leq j \leq n$) will be
worked out in the sections below.

\subsection{Segments following $\Downarrow_i$ ($1 \leq i \leq n$)}

A $\Downarrow_i$ terminator is followed by up to $i$ symbols $*$. If there
are fewer than $i$ symbols $*$, the read is finished so the segment must
be a tail. Recall that the weighted generating function of the $|$
terminator is 1, so the weighted generating function of tail segments is
$1 + z + \ldots + z^{i-1}$.

If there are $i$ symbols $*$, the sequence ends at a non-skipped position,
\textit{i.e.} in the same state as after a $\Downarrow_0$ terminator. This
is not a segment proper, because there is no terminator, but in the
transfer matrix the symbols $\Downarrow_i$ ($1 \leq i \leq n$) project
directly to the symbol $\Downarrow_0$ with weighted generating function
$z^i$.

@@

After a $\Downarrow_0$ symbol, all the counters are reset as in the
beginning of the read. If the next terminator is $|$, the segment is a
tail and we only have to make sure that it contains fewer than $\gamma$
symbols $\square$, otherwise this would create a seed. The weighted
generating function of tail segments following $\Downarrow_0$ is thus $1 +
qz + \ldots + (qz)^{\gamma-1}$.

If the next terminator is a $\downarrow_{/j}^-$ symbol ($j \geq 1$), the
segment is a match of size $j$ for the $+$ sequence. This imposes $j <
\gamma$ otherwise the segment would create a seed. Such segments consist
of $j-1$ symbols $\square$ followed by the terminator, so their weighted
generating function is $(az)^{j-1}bz$ with $1 \leq j < \gamma$.

Conversely, if the next terminator is a $\downarrow_{/j}^+$ symbol ($j
\geq 1$), we can apply the same rationale to see that the weighted
generating function is $(az)^{j-1}cz$, where $1 \leq j < \gamma$.

Finally, if the next terminator is a $\Downarrow_j$ symbol ($0 \leq j \leq
n$), we must not only make sure that the segment contains fewer than
$\gamma$ symbols $\square$, but also keep track of the position of the
next non-skipped position (stored in index $j$). Here we can follow
verbatim the rationale of section~\ref{sec:example_skip} where we replace
the probability of a match (previously $q$) by that of a double match (now
$a$), and the probability of a mismatch (previously $p$) by that a double
mismatch (now $d$). Replacing the symbols in expression (\ref{eq_H}), we
see that the weighted generating function is $(az)^x \cdot \big( 1 +
(az)^{n+1} + \ldots + (az)^{(n+1)m} \big) dz$, where  $x = n-j
\modulo{n+1}$ and $m = \left\lfloor (\gamma-1-x) / (n+1) \right\rfloor$.

\subsection{Segments following $\downarrow_{/i}^-$}

At a $\downarrow_{/i}^-$ terminator ($i \geq 1$), the read contains a
match of length $i$ for the $+$ sequence. If the next segment is a tail,
we must make sure that the match length for the $+$ sequence does not
exceed $\gamma-1$. This means that we can have up to $\gamma-i-1$ symbols
$\square$ and thus that the weighted generating function of the tail
segments is $1 + (qz) + \ldots + (qz)^{\gamma-j-1}$.

If the next terminator is a $\downarrow_{/j}^-$ symbol, we must have $j >
i$ because there is no mismatch against the $+$ sequence. In this case, we
must only make sure that the total match length for the $+$ sequence
remains lower than $\gamma$. Such segments contain $j-i-1$ symbols
$\square$ followed by the terminator so their weighted generating function
is $(qz)^{i-j-1}bz$ with $i < j \leq \gamma-1$.

For the last two types of terminators, we must pay attention to the fact
that in general, a mismatch against $+$ that follows a mismatch against
$-$ can be represented by either a $\Downarrow_j$ symbol ($0 \leq j \leq
n$) or a $\downarrow_{/j}^+$ symbol ($j \geq 1$). The terminator is
$\Downarrow_j$ if the two mismatches are within the same interval between
non-skipped positions because both sequences have match length 0. If the
mismatch against $+$ is in another interval, the terminator is
$\downarrow_{/j}^+$ because the $-$ sequence has a positive match length.

If the next terminator is $\downarrow_{/j}^+$ ($j \geq 1$), then it cannot
be in the same interval as the preceding $\downarrow_{/i}^-$ terminator,
which imposes a lower bound on the size of the segment. Since at the
$\downarrow_{/i}^-$ terminator the match length for $+$ was $i$, there
must be a non-skipped position $i$ nucleotides before. The number of
nucleotides from the $\downarrow_{/i}^-$ terminator to the next
non-skipped position is thus $y = -i \modulo{n+1}$, so the minimum segment
length is $y+1$ There is also an upper bound on the size of the segment
because the match length for the $+$ sequence cannot become higher than
$\gamma-1$, imposing the length to be lower than $\gamma-i$. The shortest
segment is terminated by $\downarrow_{/1}^+$, and the longest by
$\downarrow_{/{\gamma-y-i}}^+$. Finally, the weighted generating function
of segments terminated by $\downarrow_{/j}^+$ following the
$\downarrow_{/i}^-$ terminator is $(az)^{y+j-1}cz$ with $1 \leq j \leq
\gamma-y-i$. Note that for some value of $i$ and $y$, no $j$ can satisfy
the last inequality.

Finally, if the next terminator is $\Downarrow_j$ ($0 \leq j \leq n$) we
have to distinguish two cases. The first is that the segment length is $y$
nucleotides or shorter, where we have seen that a mismatch against $+$ is
sufficient to produce the $\Downarrow_j$ terminator. The second case is
that the segment length is above $y$ nucleotides, where a double mismatch
is necessary to produce the the $\Downarrow_j$ terminator. We have seen
above that the $\downarrow_{/i}^-$ terminator is located $y$ nucleotides
before the next non-skipped position, so in the first case, only the
terminators from $\Downarrow_0$ to $\Downarrow_{y-1}$ are possible (this
case is impossible if $y=0$). More specifically, short segment terminated
by $\Downarrow_j$ have $y-j-1$ matches followed by a mismatch for the $+$
sequence, with weighted generating function $(az)^{y-j-1}bz$.

@@

As in the case of the skip encoding, a fair amount of bookkeeping is
required to specify the weighted generating functions of interest.
We reuse the quantities $a+b+c+d=1$ and give them the same meaning as in
section~\ref{sec:dual}. For notational convenience, we define
\begin{eqnarray}
\label{eq_N}
N_i(z) &=& 1+z+\ldots+z^i, \\
W_j(z) &=& (az)^x \big( 1 + (az)^{n+1} + \ldots +
  (az)^{(n+1)m} \big) dz, \\
  \notag
  &\quad& \text{where } x = -j-1 \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma-1-x}{n+1} \right\rfloor, \\
U_{i,j}(z) &=& (az)^x
  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big)dz +
\begin{cases}
  bz\cdot(az)^x \text{ if } j < y, \\
  0 \text{ otherwise}
\end{cases} \\
  \notag
  &\quad& \text{where } x = -i-j-1 \modulo{n+1},
  y = -i \modulo{n+1}, \\
  \notag
  &\quad& \text{and } m = \left\lfloor \frac{\gamma-1-i-x}{n+1}
    \right\rfloor, \\
V_{i,j}(z) &=& (az)^x 
  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big)dz +
\begin{cases}
  cz\cdot(az)^x \text{ if } j \leq x, \\
  0 \text{ otherwise}
\end{cases} \\
  \notag
  &\quad& \text{ where } x = -i-j-1 \modulo{n+1},
  m = \left\lfloor \frac{\gamma-1-i-x}{n+1} \right\rfloor.
\end{eqnarray}
